Initial estuarine response to the nutrient dense Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-10-20 manu-draft.Rmd
seasonally corrected mult comp for wq table done

Long-term water quality monitoring data from Hillsborough and Manatee counties (Figure \@ref(fig:map)a, accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point. This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.

The historical monitoring data were also used to model an expected seasonal pattern for April to October in 2021.  This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay.  GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year. The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].  The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.  Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs provided an assessment of how 2021 data may have deviated from historical and normal seasonal variation.

Assessments were first evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication following wastewater release. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below.

Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  When appropriate, methods were used to accommodate measured concentrations in water quality variables that were below detection.  These included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].  For the multiple comparison tests, only total nitrogen, secchi, and chl-a data were evaluated, none of which were below detection as noted in the FDEP data.  Nitrate/nitrite concentrations were not evaluated quantitatively because over 80\% of the observations were below detection.

>>>>

Long-term water quality monitoring data from Hillsborough and Manatee counties (accessible at https://wateratlas.usf.edu/) were used to establish baseline conditions for the major areas of interest in Figure \@ref(fig:map)a to compare with the response monitoring data described above. Observations at each monitoring station were averaged for each month across years from 2006 to 2020. This period represents a "recovery" stage for Tampa Bay where water quality conditions were much improved from historical conditions during a more eutrophic period and when seagrass areal coverage was trending towards and above a 1950s benchmark target of 38,000 acres [@Greening14;@Sherwood17].  For each month, the mean values +/- 1 standard deviation for each parameter at each station were quantified and used as reference values relative to results at the closest monitoring station that was sampled in response to Piney Point. This comparison was made to ensure that the response data were evaluated relative to stations that were spatially relevant (e.g., long-term conditions in Terra Ceia Bay are not the same as those in middle Tampa Bay) and seasonally-specific (e.g., historical conditions in April are not the same as historical conditions in July). In some cases, the nearest long-term station did not include data for every monitoring parameter at a response location and the next closest station was used as a reference.

The historical monitoring data were also used to model an expected seasonal pattern for water quality parameters from April to October in 2021.  This was done by estimating smoothed splines using Generalized Additive Models (GAMs) using data only from the "recovery" stage of Tampa Bay (2006 to 2020).  GAMs were used to model time series of water quality parameters as a function of a continuous value for year (i.e., decimal year) and as an integer value for day of year. The continuous year value was modeled with a thin plate regression spline and the day of year value was modelled with a cyclic spline [following similar methods @Murphy19].  The modeled results provided an estimate of the expected normal seasonal variation that takes into account a long-term annual trend.  Differences in the observed values sampled in the April to October time periods from the "forecasted" predictions of the baseline GAMs through 2021 provided an assessment of how the current data may have deviated from historical and normal seasonal variation.

Statistical assessments were evaluated only on total nitrogen, chlorophyll-a, and secchi disk depth as a general analysis of potential patterns in eutrophication in nitrogen-limited systems. Observations for each data type were typically aggregated to the weekly or monthly scale given that sampling occurred at different days over the five month period.  Spatial comparisons were based primarily on the areas identified in Figure \@ref(fig:map)a. Variables with log-normal distribution were log$_10$-transformed (i.e., nutrients, chlorophyll) prior to analysis.  For statistical tests using water quality data, only the monitoring results from FDEP were used for analysis given the consistency of sample location and collection date compared to the remainder of the data obtained from other partners. Secchi observations that were visually identified on the bottom (`r sumb` observations of `r totb` in the FDEP data) were removed from analysis because these are right-censored data, whereas all other non-detects were left-censored and evaluated with methods described below.

Differences in observations between months for water quality, seagrass, and macroalgae within each area (Figure \@ref(fig:map)a) were evaluated using a Kruskal-Wallis one-way analysis of variance (ANOVA) followed by multiple comparisons using 2-sided Mann-Whitney U tests [@Hollander13]. Probability values were adjusted using the sequential Bonferroni method described in [@Holm79] to account for the increased probability of Type I error rates with multiple comparisons. An adjusted p-value < 5\% ($\alpha$ = 0.05) was considered a significant difference between months.  For water quality variables, long-term monthly averages by area were subtracted from 2021 observations to account for normal seasonal variation not attributed to potential effects from Piney Point.  Similar corrections were not done for monthly comparisons of seagrass and macroalgae data because comparable long-term seasonal data do not exist. When appropriate, methods were used to accommodate measured concentrations in water quality variables that were below detection.  These included summary statistics (e.g., median, mean, and standard deviation) following estimates of the empirical cumulative distribution functions for each parameter using the Kaplan-Meier method for censored data [@Helsel05;@Lee20].
